Racial and Ethnic Differences in ETI

### 2023-09-19 Kamyron Jordan/Racial and Ethnic Differences in ETI/racial_and_ethnic_differences_in_eti.qmd
started change point analysis

```{r setup}
#| include: false
# Libraries
library(Hmisc)
library(tidyverse)
library(gtsummary)
library(lmerTest)
library(broom.mixed)
library(emmeans)
library(knitr)
```

@@

```{r}
# Which are missing from encounters data?
a <- as.data.frame.matrix(table(annual$eDWID, annual$ReviewYear))
a <- a %>% rownames_to_column("eDWID")
e <- as.data.frame.matrix(table(by_year$eDWID, by_year$reviewyear))
e <- e %>% rownames_to_column("eDWID")
rowdiff_a <- setdiff(a, e)
rowdiff_a$data <- "annual"
rowdiff_e <- setdiff(e, a)
rowdiff_e$data <- "encounter"
diffs <- rbind(rowdiff_a, rowdiff_e)
diffs <- diffs %>%
  arrange(eDWID, desc(data)) %>%
  select(-data)
diffs <- diffs %>%
  group_by(eDWID) %>%
  reframe(across(`2015`:`2021`, ~ diff(as.numeric(.x))))
diffs %>%
  reframe(across(`2015`:`2021`, ~ sum(.x == 1))) %>%
  kable()
```

@@

```{r}
#| warning: false
annual %>%
  mutate(mean_ppFEV = as.numeric(mean_ppFEV), mean_ppFVC = as.numeric(mean_ppFVC)) %>%
  pivot_longer(c("mean_ppFEV", "mean_ppFVC")) %>%
  ggplot(aes(x = ReviewYear, y = value, color = factor(eDWID), group = eDWID)) +
  geom_line(alpha = 0.01) +
  theme_bw() +
  theme(legend.position = "none") +
  facet_wrap(~name)
```

```{r}
#| cache: true
# Most basic model of lung function over time by race
mod <- lmer(
  mean_ppFEV ~ factor(ReviewYear) * race + (1 | eDWID),
  data = annual
)
res <- tidy(mod, effects = "fixed")
means <- data.frame(emmeans(mod, ~ race + ReviewYear,
  lmerTest.limit = 1e6,
  pbkrtest.limit = 1e6
))
kable(means)
```

```{r}
ggplot(means, aes(x = ReviewYear, y = emmean, color = race)) +
  geom_line() +
  geom_point() +
  xlab("Year") +
  ylab("LS Mean FEV1") +
  geom_errorbar(aes(ymin = emmean - SE, ymax = emmean + SE), width = .2) +
  theme_bw()
```

>>>>

```{r setup}
#| include: false
# Libraries
library(Hmisc)
library(tidyverse)
library(gtsummary)
library(lmerTest)
library(nlme)
library(broom.mixed)
library(emmeans)
library(gt)
library(performance)
library(mcp)
```

@@

```{r}
# Which are missing from encounters data?
a <- as.data.frame.matrix(table(annual$eDWID, annual$ReviewYear))
a <- a %>% rownames_to_column("eDWID")
e <- as.data.frame.matrix(table(by_year$eDWID, by_year$reviewyear))
e <- e %>% rownames_to_column("eDWID")
rowdiff_a <- setdiff(a, e)
rowdiff_a$data <- "annual"
rowdiff_e <- setdiff(e, a)
rowdiff_e$data <- "encounter"
diffs <- rbind(rowdiff_a, rowdiff_e)
diffs <- diffs %>%
  arrange(eDWID, desc(data)) %>%
  select(-data)
diffs <- diffs %>%
  group_by(eDWID) %>%
  reframe(across(`2015`:`2021`, ~ diff(as.numeric(.x))))
diffs %>%
  reframe(across(`2015`:`2021`, ~ sum(.x == 1))) %>%
  gt()
```

@@

### FEV1

#### Basic model

The most basic model we can fit to these data uses the average values as the outcome, with terms for review year, race, and their interaction and a random intercept for each patient.

```{r}
#| cache: true
# Most basic model of lung function over time by race
# Try a few different models
mod <- lmer(
  mean_ppFEV ~ factor(ReviewYear) * race + (1 | eDWID),
  data = annual
)
res <- tidy(mod, effects = "fixed")
means <- data.frame(emmeans(mod, ~ race + ReviewYear,
  lmerTest.limit = 1e6,
  pbkrtest.limit = 1e6
))
label(means) = list("Race/Ethnicity","Year","Mean ppFEV","SE","df","Lower CL", "Upper CL")
gt(means) %>% fmt_number(emmean:last_col(),decimals = 3)
```

```{r}
ggplot(left_join(annual, means, by = join_by(race, ReviewYear))) +
  geom_line(mapping = aes(x = ReviewYear, y = mean_ppFEV, group = eDWID),alpha=0.005,linewidth=0.5) +
  geom_point(mapping = aes(x = ReviewYear, y = emmean,color=race)) +
  geom_line(mapping = aes(x = ReviewYear, y = emmean,group = race,color=race)) +
  geom_errorbar(mapping=aes(x = ReviewYear,ymin = emmean - SE, ymax = emmean + SE,color=race), width = .2) +
  xlab("Year") +
  ylab("LS Mean FEV1") +
  theme_bw() +
  scale_color_discrete(name = "Race/Ethnicity")
```

```{r}
ggplot(left_join(annual, means, by = join_by(race, ReviewYear))) +
  geom_line(mapping = aes(x = ReviewYear, y = mean_ppFEV, group = eDWID),alpha=0.005,linewidth=0.5) +
  geom_point(mapping = aes(x = ReviewYear, y = emmean,color=race)) +
  geom_line(mapping = aes(x = ReviewYear, y = emmean,group = race,color=race)) +
  geom_errorbar(mapping=aes(x = ReviewYear,ymin = emmean - SE, ymax = emmean + SE,color=race), width = .2) +
  xlab("Year") +
  ylab("LS Mean FEV1") +
  theme_bw() +
  scale_color_discrete(name = "Race/Ethnicity") +
  facet_wrap(~race)
```

#### Change point model

```{r}
#| eval: false
t = annual %>% select(eDWID,ReviewYear,mean_ppFEV) %>% na.omit(.)
model = list(mean_ppFEV ~ 0 + ReviewYear, 
              ~ 0 + ReviewYear)
fit = mcp(model,data = t,cores=4,chains = 4)
plot(fit)
```
